Principled Drug-Drug Interaction Terms for Bayesian Logistic Regression Models of Drug Safety in Oncology Phase I Combination Trials

### 2024-10-04 manuscript.tex

\usepackage{pifont}% http://ctan.org/pkg/pifont
\usepackage{makecell}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

\usepackage{subcaption}

\captionsetup[subfigure]{labelfont={bf,sf}, labelformat=simple, labelsep=none, textfont=normalfont, font=normalsize, singlelinecheck=off, justification=raggedright} %
\renewcommand{\thesubfigure}{\Alph{subfigure}}

@@

% \footnotetext{\textbf{Abbreviations:} BLRM, Bayes; APC, antigen-presenting cells; IRF, interferon regulatory factor}

\newpage
\section{Introduction}\label{sec1}
In Phase-I Oncology, dose escalation trials are concerned with striking the right balance of ensuring patient safety while rapidly escalating drug doses to avoid exposure of patients to sub-therapeutic dosing and determining the maximum tolerated dose\cite{LeTourneau2009}. This is motivated by the cytotoxic nature of (older) cancer drugs, where both efficacy and toxicity were expected to increase with dose. In contrast to drugs targeting other diseases, lack of efficacy leads to death, and as a consequence, some level of toxicity is acceptable. 
The goals of such dose-escalation studies are to systematically increase dosing
\begin{enumerate}
\item as quickly as possible, such that a biologically active (and potentially efficacious) dose is reached as soon as possible, and to keep overall study size small,
\item to determine the recommended phase II dose (RP2D), for instance, via the maximum tolerated dose (MTD), for use in future trials\cite{Neuenschwander2014},
\item while doing so as safely as possible, i.e., while controlling the rate of dose-limiting toxicity (DLT) events during the trial, as well as controlling the probability of declaring an overly toxic final dose (which would be a risk to any downstream trials).
\end{enumerate}
Typically, such escalation is performed in treatment cycles (e.g. of length 4 weeks), where patients are treated according to a defined schedule. DLT events are recorded for the duration of the treatment cycle, and based on these observations a dose for the next cohort of patients is chosen. Combination therapies have become a quite common strategy to (among other considerations) increase efficacy \cite{Mokhtari2017, Palmer2017} while maintaining an acceptable safety profile. Thus, we may also want to explore the safety of the compound of interest in combination with other compounds (such as an established standard of care), necessitating modelling safety-relevant drug-drug interactions on top of the single drug-toxicity relationships.

@@

In modern Oncology, we apply model-based approaches to more efficiently conduct trials. Both for single-drug and combinations, these approaches model the dose-toxicity relationship in order to limit the probability of administering an overdose, for instance using the EWOC criterion. Since combination therapies have become rather common, the ability of any model-based approach to properly account for drug-drug interactions of those drug combinations has become increasingly important. Two drugs can exhibit synergistic toxicity, i.e., combining them leads to higher toxicity than expected from the two drugs individually, or antagonistic toxicity, i.e., combining them leads to lower toxicity than expected from the two drugs individually (we will formalize this later on). In the case of the BLRM, two and three-drug interaction models have been developed \cite{Neuenschwander2014} and more recently been extended to $N$ drugs \cite{OncoBayes2}. Other popular single-drug safety models, such as the Continual Reassessment Method (CRM)\cite{Quigley1990} or Bayesian optimal interval design (BOIN)\cite{Yuan2016}, also have more recent extensions for modelling drug combinations (e.g., the POCRM\cite{Wages2011} or BOIN-COMB\cite{Zhang2016, Lin2017}).

Here, we will detail a principled drug-drug interaction model for the BLRM method. This model is now implemented as the new default drug-drug interaction model (from version 0.8-0 onwards) in the open-source OncoBayes2\cite{OncoBayes2} R package available on CRAN. Note that the focus of this work is on principles and illustrative examples using data scenarios, and explicitly \emph{not} to compare the principled model against others in operating characteristics simulation studies, since the outcome of such simulations depends heavily on the design of the particular trial, scenarios chosen, and comparing methods that are applicable. We provide code to run the models on the data scenarios detailed in this manuscript as well as produce the figures using the OncoBayes2 R package, which can be re-used to implement custom scenarios or simulations relevant to the trial design of interest. Thus, our focus is on deriving an improved drug-drug interaction model by formalizing desirable properties and designing a drug-drug interaction model for the BLRM that satisfies them. In principle, these properties can also be used as guidance to design drug-drug interaction models for other methods.

@@

    \item If any drug dose goes to infinity, the probability of a DLT should approach one, regardless of the other drug doses and  interaction strengths. This property formalizes the intuitive concept of "the dose makes the poison", credited to Paracelsus \cite{Grandjean2016}. I.e., at high doses every substance will become toxic -- this even holds for water \cite{ROWNTREE1923, Gardner2002}. For example, the two-drug model should fulfill

@@

    \item Last, but not least, for drugs that exhibit antagonistic toxicity, a desired property is to be able to model locally non-monotonic dose-responses. A classical example here would be treatment of methanol intoxication by administering ethanol: methanol and ethanol in combination are less toxic than methanol alone with respect to inducing blindness\cite{Marraffa2012} (at least up to a point; see property \eqref{eq:asymptoticToxicity}). More formally, for any finite choice of $d_1 > 0, d_2 > 0$ and $m_1>0, m_2 > 0$ with $0<\epsilon_1 <m_1, 0<\epsilon_2<m_2$, a choice of parameters exists such that toxicity is locally decreasing and we can find

@@

    More generally, if the drugs in set $s$ exhibit a synergistic or antagonistic toxicity, while all other drug combinations do not interact / act independently, we would like to enforce
    \begin{IEEEeqnarray}{rCl}
    	 \pi_{\bm{\eta}}(\bm{d}) & > & \pi_{\perp}(\bm{d}) \text{ where } \eta_s > 0, \text{ or }  \label{eq:synergisticToxicity} \\
    	 \pi_{\bm{\eta}}(\bm{d}) & < & \pi_{\perp}(\bm{d}) \text{ where } \eta_s < 0, \text{ respectively, with} \label{eq:antagonisticToxicity} \\
    	 \bm{\eta} & = &  \left(0, \dots, 0, \eta_s, 0, \dots, 0\right)^\intercal, s \in \mathcal I_N. \IEEEnonumber
    \end{IEEEeqnarray}
    In addition, for antagonistically-toxic cases, we require the model be locally non-monotonic (again, if all other drug combinations interact independently). That is, for any finite choice of $ d_{\bm{s}} > \bm{0}, \bm{m}>\bm{0}, 0<\epsilon_i<m_i, i \in s$,  a choice of parameters exists such that toxicity is locally decreasing and we can find
    \begin{IEEEeqnarray}{rCl}
	    \pi_{\bm{\eta}}(\bm{d_{s}} + \bm{\epsilon_s}, \bm{d_{s^\complement}}) & < & \pi_{\bm{\eta}}(\bm{d_{s}}, \bm{d_{s^\complement}}) \label{eq:antagonisticToxicityNonMonotonic}
    \end{IEEEeqnarray}
\end{enumerate}

@@

\section{Discussion}\label{sec:discussion}
Drug-drug interactions in Oncology have received substantial attention over the last two decades \cite{Beijnen2004}, and ways of systematically addressing them when modelling dose-dependent DLT probability in phase-I Oncology have been established \cite{Thall2003}. In contrast to the recently-proposed notion that ignoring drug-drug interactions when modelling safety in phase-I Oncology combination trials is beneficial in general \cite{Mozgunov2021}, we think that this depends on the type of interaction model used, and how it influences the dose-toxicity surface. Furthermore, if (pre-)clinical data on drug-drug interactions are available (for instance, through drug metabolism), these can be incorporated in the interaction prior, for instance  by defining a prior for the odds multiplier at a specific dose. Note that one could also try applying more mechanistic drug interaction models with more parameters as summarized in a recent review\cite{Meyer2020} (especially if strong evidence for a particular mechanism is available), however, in the context of general Phase I Oncology dose escalation trials, the identifiability of such models would be questionable given the sparse data typically available.

According to the analytical criteria in section \ref{sec:methodology}, the presented drug-drug interaction model for the BLRM is an improvement over the previous state-of-the art linear model in logit space, as well as earlier models\citep{Thall2003}. It prevents the probability of toxicity to go to 0 when the dose of a modelled drug goes to infinity, but still supports non-monotonic combination dose-responses in scenarios with antagonistic toxicity. It could also enable running joint models for platform trials where arms with $N - 1$ drugs and $N$ drugs run simultaneously without jeopardizing accurate estimation of the marginals for the individual drug components if a suitable prior is chosen. This is now the default drug-drug interaction model in the OncoBayes2 R package from version 0.8-0 onwards.

Last, but not least, our hope is that the improved interaction model will facilitate the use of joint models, which could also make more efficient use of the data since the model structure is more appropriate for the problem at hand. The previously uncovered flaws in older interaction models prevented their use in joint models in practice, or led to them \emph{de-facto} being disabled through use of tight priors on the interaction strength parameters $\eta$ in practice. Given the improved interaction model, one can now be more sensitive to potential drug-drug interactions without risking undesired behavior at higher doses.

@@

\section*{Acknowledgments}
We would like to thank Beat Neuenschwander and Matt Whiley for feedback on the proposed model and structure of the manuscript.

@@

Lukas Widmer and Sebastian Weber are employed by Novartis Pharma AG and own stocks in the company. Andrew Bean and David Ohlssen are employed by Novartis Pharmaceuticals Corporation and own stocks in the company.

>>>>

\usepackage{pifont}% http://ctan.org/pkg/pifont
\usepackage{makecell}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\revision}[1]{\textcolor{blue}{#1}}
\usepackage{subcaption}

\captionsetup[subfigure]{labelfont={bf,sf}, labelformat=simple, labelsep=none, textfont=normalfont, font=normalsize, singlelinecheck=off, justification=raggedright} %
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}

@@

\newpage
\section{Introduction}\label{sec1}
In Phase-I Oncology, dose escalation trials are concerned with striking the right balance of ensuring patient safety while rapidly escalating drug doses to avoid exposure of patients to sub-therapeutic dosing and determining the maximum tolerated dose\cite{LeTourneau2009}. This is motivated by the cytotoxic nature of (older) cancer drugs, where both efficacy and toxicity were expected to increase with dose. In contrast to drugs targeting other diseases, lack of efficacy leads to death, and as a consequence, some level of toxicity is acceptable. 
The goals of such dose-escalation studies are to systematically increase dosing
\begin{enumerate}
\item as quickly as possible, such that a biologically active (and potentially efficacious) dose is reached as soon as possible, and to keep overall study size small,
\item to determine the recommended phase II dose (RP2D), for instance, via the maximum tolerated dose (MTD\footnote{\revision{Note that for combination therapies, the MTD is not a unique dose combination, but a maximally tolerated dose combination surface, and efficacy is likely to be different at different points on this surface.}}), for use in future trials\cite{Neuenschwander2014},
\item while doing so as safely as possible, i.e., while controlling the rate of dose-limiting toxicity (DLT) events during the trial, as well as controlling the probability of declaring an overly toxic final dose (which would be a risk to any downstream trials).
\end{enumerate}
Typically, such escalation is performed in treatment cycles (e.g. of length 4 weeks), where patients are treated according to a defined schedule. DLT events are recorded for the duration of the treatment cycle, and based on these observations a dose for the next cohort of patients is chosen. Combination therapies have become a quite common strategy to (among other considerations) increase efficacy \cite{Mokhtari2017, Palmer2017} while maintaining an acceptable safety profile. Thus, we may also want to explore the safety of the compound of interest in combination with other compounds (such as an established standard of care), necessitating modelling safety-relevant drug-drug interactions on top of the single drug-toxicity relationships.

@@

In modern Oncology, we apply model-based approaches to more efficiently conduct trials. Both for single-drug and combinations, these approaches model the dose-toxicity relationship in order to limit the probability of administering an overdose, for instance using the EWOC criterion. As combination therapies have become rather common\revision{, early methods that extend the BLRM to two-drug combination settings have been developed already 20 years ago, e.g., the 6-parameter combination BLRM \cite{Thall2003}. Further extensions have then been made since, for instance, 4-parameter BLRMs for univariate dose escalation in each escalation step / cohort with discrete doses \cite{Shi2013}, or in continuous dose space to estimate the maximum tolerated dose curve\cite{Tighiouart2014}, as well as to explore dose space in both dimensions in each cohort\cite{Tighiouart2017}. With the rise of combination therapies and higher-order combinations (such as a recent quadruple combination therapy in metastatic pancreatic ductal adenocarcinoma\cite{Wainberg2023}), the BLRM methodology was evolved further, e.g., with a triple-combination BLRM variant \cite{Tighiouart2016}, and} the ability of any model-based approach to properly account for drug-drug interactions of such drug combinations has also become increasingly important.

Already two drugs can exhibit synergistic toxicity, i.e., combining them leads to higher toxicity than expected from the two drugs individually, or antagonistic toxicity, i.e., combining them leads to lower toxicity than expected from the two drugs individually (we will formalize this later on). \revision{In particular, antagonistic toxicity can lead to non-monotonic dose-response curves / surfaces}. In the case of the BLRM, two and three-drug interaction models have been developed \cite{Neuenschwander2014} and more recently been extended to $N$ drugs \cite{OncoBayes2}. Other popular single-drug safety models, such as the Continual Reassessment Method (CRM)\cite{Quigley1990} or the Bayesian optimal interval design (BOIN)\cite{Yuan2016}, also have more recent extensions for modelling drug combinations (e.g., the POCRM\cite{Wages2011} or BOIN-COMB\cite{Zhang2016, Lin2017})\revision{, and there are more recently developed methods specifically designed for combination trials, e.g., the PIPE\cite{Mander2015} and NBCD\cite{Razaee2022} methods. Using copulas for modeling multi-drug DLT probabilities has also been suggested\cite{Yin2009}. Notably, while all of the cited methods consider drug-drug interactions implicitly or explicitly, they also all assume a partial order for the dose-toxicity relationship, even under antagonistic drug-drug interactions, which may not always hold. In particular, neglecting this aspect may lead to sub-optimal dosing for combination therapies from a benefit/risk aspect, as increasing the dose of one or both of the drugs may lead to lower toxicity and may simultaneously also affect efficacy.}

\revision{Practical examples for two drugs exhibiting an interaction in terms of their toxicity include altering each other's metabolism, for instance through Cytochrome P450 enzymes\cite{Tanaka1998}. E.g., if one or both of the drugs in question inhibit an enzyme that is involved in their metabolism, a synergistic drug-drug toxicity interaction might be expected, as the drugs will get metabolized more slowly and accumulate (note that we do not consider active metabolites of drugs here). Conversely, if one or both drugs induce such an enzyme, an antagonistic drug-drug toxicity interaction might be expected. Notably, in such cases, increasing one of the drug doses may decrease overall toxicity, leading to non-monotonicity of the dose-response. Another classical example of non-monotonic dose-response in terms of toxicity would be treatment of methanol intoxication by administering ethanol: methanol and ethanol in combination are less toxic than methanol alone with respect to inducing blindness\cite{Marraffa2012}, at least up to a point. Obviously, at very high doses, every substance will become toxic, i.e., more intuitively, "the dose makes the poison" (credited to Paracelsus\cite{Grandjean2016}) -- this even holds for water \cite{ROWNTREE1923, Gardner2002}.}

In this work, we will develop a principled drug-drug interaction model \revision{for arbitrary-order drug combination therapies using the BLRM method}. This model is now implemented as the new default drug-drug interaction model (from version 0.8-0 onwards) in the open-source OncoBayes2\cite{OncoBayes2} R package available on CRAN. Note that the focus of this work is on principles and illustrative examples using data scenarios, and explicitly \emph{not} to compare the principled model against others in operating characteristics simulation studies, since the outcome of such simulations depends heavily on the design of the particular trial, scenarios chosen, and comparing methods that are applicable, \revision{especially as most dose escalation methods do not consider non-monotonic dose-responses -- though the design principles for drug-drug interaction terms outlined here could likely be applied to methods other than the BLRM. For easy reproducibility,} we provide code to run the models on the data scenarios detailed in this manuscript as well as produce the figures using the OncoBayes2 R package, which can be re-used to implement custom scenarios or simulations relevant to the trial design of interest. Thus, our focus is on deriving an improved drug-drug interaction model by formalizing desirable properties and designing a drug-drug interaction model for the BLRM that satisfies the properties outlined below. In principle, these properties can also be used as guidance to design drug-drug interaction models for other methods.

@@

    \item If any drug dose goes to infinity, the probability of a DLT should approach one, regardless of the other drug doses and  interaction strengths. This property formalizes the intuitive concept of "the dose makes the poison" mentioned in the introduction. For example, the two-drug model should fulfill

@@

    \item Last, but not least, for drugs that exhibit antagonistic toxicity, a desired property is to be able to model locally non-monotonic dose-responses. \revision{More formally, at a finite choice of $d_1 > 0, d_2 > 0$, and $\epsilon_1, \epsilon_2 \in \Re_{\geq 0}, \epsilon_1 + \epsilon_2 > 0$, a choice of parameters exists such that toxicity is locally decreasing, i.e.,

@@

    \end{IEEEeqnarray*}}

    More generally, if the drugs in set $s$ exhibit a synergistic or antagonistic toxicity, while all other drug combinations do not interact / act independently, we would like to enforce
    \begin{IEEEeqnarray}{rCl}
    	 \pi_{\bm{\eta}}(\bm{d}) & > & \pi_{\perp}(\bm{d}) \text{ where } \eta_s > 0, \text{ or }  \label{eq:synergisticToxicity} \\
    	 \pi_{\bm{\eta}}(\bm{d}) & < & \pi_{\perp}(\bm{d}) \text{ where } \eta_s < 0, \text{ respectively, with} \label{eq:antagonisticToxicity} \\
    	 \bm{\eta} & = &  \left(0, \dots, 0, \eta_s, 0, \dots, 0\right)^\intercal, s \in \mathcal I_N. \IEEEnonumber
    \end{IEEEeqnarray}
    \revision{
    In addition, for antagonistically-toxic cases, we require the model be specified in a way that allows for local non-monotonicity (again, if all other drug combinations interact independently). That is, for any finite choice of $\bm{d_{s}} > \bm{0}, \epsilon_i \in \Re_{\geq 0}, i \in s, \sum_{i \in s} \epsilon_i > 0$,  a choice of parameters exists such that toxicity is locally decreasing, i.e.,
    \begin{IEEEeqnarray}{rCl}
	    \pi_{\bm{\eta}}(\bm{d_{s}} + \bm{\epsilon_s}, \bm{d_{s^\complement}}) & < & \pi_{\bm{\eta}}(\bm{d_{s}}, \bm{d_{s^\complement}}) \label{eq:antagonisticToxicityNonMonotonic}
    \end{IEEEeqnarray}}
\end{enumerate}

@@

\section{Discussion}\label{sec:discussion}
Drug-drug interactions in Oncology have received substantial attention over the last two decades \cite{Beijnen2004}, and ways of systematically addressing them when modelling dose-dependent DLT probability in phase-I Oncology have been established \cite{Thall2003}. \revision{Concomitantly, there has been discussion around whether to model such interactions at all and data requirements to identify them. For instance, \cite{Wang2005, Riviere2015} recommended against using an interaction term, since the interaction parameter(s) cannot reliably be estimated using typically small sample sizes of phase I trials, and \cite{Mozgunov2021} supported those statements with extensive simulations. However, Tighiouart \emph{et al.}\cite{Tighiouart2022} contrasted those claims when estimating either the MTD contour or a single MTD, where safety of the trial may be compromised if a highly significant interaction term is omitted from the model. In our experience, this is indeed be an issue to be taken seriously, especially when constructing combination models where single-drug / lower-order combination historical data are included and pooled with the combination data (i.e., in the absence of a hierarchical model for between-trial-heterogeneity). In such cases, safety may be compromised both during the trial and for MTD selection, as illustrated in the example data scenario in Figure~\ref{fig:prior_histdata_5_5_100_05}, where the no-interaction model still allows the dose combination of 200/200, even though 5 patients with 5 DLTs have been observed at that dose -- i.e., in the absence of a drug-drug interaction term, the single-drug data overrules the combination data, especially when there is only little / initial combination data.}

Additionally, if (pre-)clinical data on drug-drug interactions are available (for instance, through drug metabolism), these can be incorporated in the interaction prior, for instance by defining a prior for the odds multiplier at a specific dose. Note that one could also try applying more mechanistic drug interaction models with more parameters as summarized in a recent review\cite{Meyer2020} (especially if strong evidence for a particular mechanism is available), however, in the context of general Phase I Oncology dose escalation trials, the identifiability of such models would be questionable given the sparse data typically available.

\revision{The presented revised drug-drug interaction model for the BLRM improves over the previous state-of-the art linear model in logit space, as well as an earlier model\citep{Thall2003} in terms of fulfilling the desired properties for modeling such interactions outlined in section \ref{sec:methodology}.} It prevents the probability of toxicity to go to 0 when the dose of a modelled drug goes to infinity, but still supports non-monotonic combination dose-responses in scenarios with antagonistic toxicity. It could also enable running joint models for platform trials where arms with $N - 1$ drugs and $N$ drugs run simultaneously without jeopardizing accurate estimation of the marginals for the individual drug components if a suitable prior is chosen. This drug-drug interaction model is been implemented as the default in the OncoBayes2 R package from version 0.8-0 onwards.

Last, but not least, our hope is that the improved interaction model will facilitate the use of joint models, which could also make more efficient use of the data since the model structure is more appropriate for the problem at hand. The previously uncovered flaws in some of the older interaction models prevented their use in joint models in practice, or led to them \emph{de-facto} being disabled through use of tight priors on the interaction strength parameters $\eta$ \revision{in BLRMs} in practice. Given this improved interaction model, one can now be more sensitive to potential drug-drug interactions without risking undesired behavior at higher doses.

@@

\section*{Acknowledgments}
We would like to thank Beat Neuenschwander, Matt Whiley\revision{, Nikolas Siccha, as well as the anonymous reviewers} for feedback on the proposed model and structure of the manuscript.

@@

Lukas Widmer and Sebastian Weber are employed by Novartis Pharma AG and own stocks in the company. Andrew Bean and David Ohlssen are employed by Novartis Pharmaceuticals Corporation and owns stocks in the company.
